---
title: "buildwright cookbook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{buildwright cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(buildwright)
```

This cookbook collects end-to-end recipes for the kind of problem buildwright
exists to catch: a managed R distribution (Posit Connect / Workbench,
Bioconductor, an internal package repository) that *looks* fine in a lockfile
but will not actually install. Every recipe here runs offline against bundled
or locally-constructed data, so you can reproduce each one without a network
connection. The fuller, network-enabled versions live in the package's
`cookbook/` folder (run with `Rscript cookbook/01-...R` from the package root).

A note on list columns: `requirements`, `required_by`, and `dependents` are
list columns, which `knitr::kable()` and `DT` cannot print directly. Where a
recipe shows such a column in a table, it is flattened to a comma-joined string
first with `vapply(x, paste, character(1), collapse = ", ")`.

## Recipe 1: Diagnosing a deliberately-broken Bioconductor install

A common Bioconductor failure mode: a lockfile pins a Bioconductor package but
omits one of its Bioconductor dependencies (for example because the dependency
was only present transitively in the environment the lock was captured from).
The install then dies part-way through with an opaque "package 'X' is not
available" error. buildwright surfaces this *before* the install runs.

We build a tiny broken `renv.lock` inline, write it to a temp file, and scan
it. `S4Vectors` requires `IRanges`, but `IRanges` is absent from the lockfile.
We also include `RCurl`, which carries a known system requirement (libcurl), so
the same diagnosis demonstrates a system-library warning.

```{r build-broken-lock}
broken_lock <- '{
  "R": { "Version": "4.3.2" },
  "Bioconductor": { "Version": "3.18" },
  "Packages": {
    "BiocGenerics": {
      "Package": "BiocGenerics",
      "Version": "0.48.1",
      "Source": "Bioconductor",
      "Repository": "Bioconductor 3.18",
      "Requirements": []
    },
    "S4Vectors": {
      "Package": "S4Vectors",
      "Version": "0.40.2",
      "Source": "Bioconductor",
      "Repository": "Bioconductor 3.18",
      "Requirements": ["BiocGenerics", "IRanges"]
    },
    "RCurl": {
      "Package": "RCurl",
      "Version": "1.98-1.14",
      "Source": "Repository",
      "Repository": "CRAN",
      "Requirements": ["bitops"]
    },
    "bitops": {
      "Package": "bitops",
      "Version": "1.0-7",
      "Source": "Repository",
      "Repository": "CRAN",
      "Requirements": []
    }
  }
}'

lock_path <- tempfile("broken-bioc-", fileext = ".lock")
writeLines(broken_lock, lock_path)
```

Scan it and confirm the sources are read correctly (two Bioconductor, two
CRAN):

```{r scan-broken}
lib <- scan_library(lock_path)
lib[, c("package", "version", "source")]
table(lib$source)
```

`detect_conflicts()` flags the missing Bioconductor dependency. `IRanges` is
required by `S4Vectors` but is absent from the scanned set, so it is reported as
`missing` and unsatisfiable:

```{r conflicts-broken}
conf <- detect_conflicts(lib)
conf[, c("package", "type", "satisfiable", "detail")]
```

`simulate_install()` then predicts the consequence: `IRanges` has no install
step (it is not in the set), and everything that depends on it transitively is
`blocked`, so the plan is not feasible:

```{r simulate-broken}
plan <- simulate_install(lock_path, platform = "linux", host_libraries = character())
plan$feasible
plan$predicted_failures[, c("package", "predicted_status", "reason")]
```

Finally, `check_sysreqs()` (on a host with no system libraries) catches the
libcurl requirement that `RCurl` pulls in:

```{r sysreqs-broken}
sr <- check_sysreqs(lib, platform = "linux", host_libraries = character())
sr[, c("package", "system_requirement", "status", "install_hint")]
```

A single `diagnose()` call bundles all of the above; its print method gives the
at-a-glance health summary you would gate an install on:

```{r diagnose-broken}
diag <- diagnose(lock_path, platform = "linux", host_libraries = character())
diag
summary(diag)
```

## Recipe 2: Auditing a real `renv.lock` from a public R project

buildwright ships a real lockfile taken from the `rstudio/renv` test resources
(`tests/testthat/resources/bioconductor.lock`). It is a genuine, if small,
Bioconductor environment and contains two real reproducibility smells worth
catching in review.

```{r load-public}
real_lock <- system.file("extdata", "public-project.lock", package = "buildwright")
rlib <- scan_library(real_lock)
rlib[, c("package", "version", "source", "repository")]
```

The source breakdown shows the first finding immediately: alongside two
Bioconductor packages there is a package installed from a **Local** source.

```{r public-sources}
table(rlib$source)
```

**Finding 1 -- a Local-source package.** `renv` itself is recorded with
`Source: "Local"` pointing at a path on the lock author's machine
(`~/r/pkg/renv`). A Local source is not reproducible on any other machine: there
is no repository or commit to restore from. In a review you would flag this and
ask for a CRAN or pinned-GitHub version instead.

```{r public-local}
rlib[rlib$source == "local", c("package", "version", "repository")]
```

**Finding 2 -- a Bioconductor release mismatch.** Both Bioconductor packages
report the same `source`, but the lockfile pins them to *different Bioconductor
release branches*. The `git_branch` fields are not surfaced in the scan tibble,
so we read the raw lock JSON with `jsonlite` to expose them:

```{r public-bioc-mismatch}
raw <- jsonlite::fromJSON(real_lock, simplifyVector = FALSE)
bioc <- Filter(function(p) identical(p$Source, "Bioconductor"), raw$Packages)
branch_of <- function(p) if (is.null(p$git_branch)) NA_character_ else p$git_branch
data.frame(
  package = vapply(bioc, function(p) p$Package, character(1)),
  version = vapply(bioc, function(p) p$Version, character(1)),
  git_branch = vapply(bioc, branch_of, character(1)),
  row.names = NULL
)
```

`BiocGenerics` is pinned to `RELEASE_3_13` while `limma` is pinned to
`RELEASE_3_14` -- two different Bioconductor releases in one environment. The
lockfile's declared `Bioconductor$Version` is `3.13`, so the `limma` pin is
inconsistent with the environment it claims to describe. That is exactly the
sort of drift that makes a Bioconductor restore fail or silently mix releases.

The structural diagnosis itself is clean (these three packages have no missing
dependencies or cycles), which is the point: a lockfile can pass the structural
checks and still carry reproducibility problems a human should see.

```{r public-diagnose}
diagnose(real_lock)
```

If you wanted to fetch this lock live from GitHub instead of using the bundled
copy, you would do the following (not run here, since the vignette must build
offline):

```{r public-download, eval = FALSE}
url <- paste0(
  "https://raw.githubusercontent.com/rstudio/renv/main/",
  "tests/testthat/resources/bioconductor.lock"
)
dest <- tempfile("public-", fileext = ".lock")
download.file(url, dest, quiet = TRUE)
diagnose(dest)
```

## Recipe 3: Catching a version-drift conflict between two CRAN packages

Lockfiles often omit version constraints, but installed libraries (and many
internal package sets) carry them in each `DESCRIPTION`. When two packages
demand incompatible version ranges of a shared dependency, no single version
can satisfy both and the install is doomed. buildwright detects this from the
`DESCRIPTION` files directly.

We construct a tiny three-package library in a temp directory: `pkgA` imports
`shared (>= 2.0)`, `pkgB` imports `shared (<= 1.0)`, and `shared` is present at
`1.5`. The two ranges do not intersect.

```{r build-drift-lib}
lib_dir <- file.path(tempdir(), "drift-lib")
dir.create(lib_dir, showWarnings = FALSE)

write_desc <- function(name, version, imports = NULL) {
  pkg_dir <- file.path(lib_dir, name)
  dir.create(pkg_dir, showWarnings = FALSE)
  lines <- c(
    paste0("Package: ", name),
    paste0("Version: ", version),
    "Title: Fixture",
    "Description: Fixture package for the buildwright cookbook.",
    "License: GPL-3"
  )
  if (!is.null(imports)) lines <- c(lines, paste0("Imports: ", imports))
  writeLines(lines, file.path(pkg_dir, "DESCRIPTION"))
}

write_desc("pkgA", "1.0.0", imports = "shared (>= 2.0)")
write_desc("pkgB", "1.0.0", imports = "shared (<= 1.0)")
write_desc("shared", "1.5.0")
```

Scan the directory as an installed library and detect conflicts. The shared
dependency is flagged as a `version_conflict`: the intersection of `>= 2.0` and
`<= 1.0` is empty.

```{r drift-conflicts}
drift_lib <- scan_library(lib_dir)
drift_lib[, c("package", "version", "source", "requirements")]

dconf <- detect_conflicts(drift_lib)
dconf[, c("package", "type", "constraints", "installed_version", "satisfiable")]
```

The `constraints` column records both demands so the reviewer can see exactly
which packages disagree:

```{r drift-detail}
dconf$detail
```

`simulate_install()` marks `shared` itself as failing (unsatisfiable
constraints) and `pkgA`/`pkgB` as blocked behind it:

```{r drift-simulate}
dplan <- simulate_install(drift_lib)
dplan$feasible
dplan$predicted_failures[, c("package", "predicted_status", "reason")]
```

## Recipe 4: Wiring buildwright into renv / pak

The point of all of the above is to fail *fast and early*, before an install
spends ten minutes compiling and then dies. `bw_preinstall_check()` is the gate:
run it in front of `renv::restore()` or a `pak` install and, with `strict =
TRUE`, it aborts when problems are predicted.

On a healthy lockfile the strict check passes silently and the install
proceeds:

```{r hook-clean}
bw_preinstall_check(bw_example("clean.lock"), strict = TRUE, quiet = TRUE)
```

On a broken lockfile, `strict = TRUE` raises an error -- in a script or CI job
that aborts before any package is built. Here we catch the error to show its
message:

```{r hook-broken}
tryCatch(
  bw_preinstall_check(bw_example("conflicted.lock"), strict = TRUE, quiet = TRUE),
  error = function(e) cat("gate tripped:", conditionMessage(e), "\n")
)
```

In a real restore script you would place the gate immediately before the
install call (shown here with `eval = FALSE` because it touches the network and
your library):

```{r hook-renv, eval = FALSE}
# restore.R -- run before bringing an environment up to its lockfile.
buildwright::bw_preinstall_check("renv.lock", strict = TRUE)
renv::restore()
```

The same gate works in front of `pak`. On a host where you know which system
libraries are installed, pass them so the system-requirement check is
deterministic:

```{r hook-pak, eval = FALSE}
buildwright::bw_preinstall_check(
  "renv.lock",
  strict = TRUE,
  host_libraries = c("libcurl", "openssl")
)
pak::pkg_install("local::.")
```

And as a CI step, a one-line `Rscript` invocation fails the job (non-zero exit)
when the lockfile is unhealthy:

```sh
Rscript -e 'buildwright::bw_preinstall_check("renv.lock", strict = TRUE)'
```

See the package's `cookbook/` folder for the full, network-enabled versions of
recipes 1-2 (rendering HTML reports and downloading a live lockfile) and a
ready-to-drop-in `cookbook/hooks/preinstall-hook.R`.
