---
title: "Getting started with buildwright"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with buildwright}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(buildwright)
```

## The problem

Platform and DevOps teams that maintain a shared R distribution rarely fail an
install because of a typo. They fail because a few hundred packages, pinned in an
`renv.lock` and spread across CRAN, Bioconductor and GitHub, quietly disagree:
one package needs `rlang (>= 1.1.0)` while another caps it at `<= 0.4.12`, a
required package was never recorded, a pair of packages depend on each other in a
cycle, or a binary needs a system library (`libcurl`, `GDAL`, `OpenSSL`) that the
build image does not ship. buildwright reads the lockfile *before* you build,
models the whole graph, and tells you exactly which packages will fail and why --
so the dry run happens in milliseconds instead of after twenty minutes of
compilation on a fresh machine.

Everything below runs against lockfiles bundled with the package, reachable
through `bw_example()`, so no network or installed library is required.

## Scan a lockfile

`scan_library()` auto-detects its input -- here, an `renv.lock` path -- and
returns a tidy `bw_library` tibble with one row per package.

```{r scan}
lib <- scan_library(bw_example("clean.lock"))
lib
```

Each package carries a normalised `source`. The clean fixture spans three of
them:

```{r scan-source}
table(lib$source)
```

## Build the dependency graph

`build_dep_graph()` turns the library into a directed `igraph` where an edge
`A -> B` means "A requires B". From it you can ask for cycles and for diamond
(shared) dependencies -- the packages that multiple others depend on, which are
exactly where version conflicts tend to arise.

```{r graph}
g <- build_dep_graph(lib)
g

bw_cycles(g)
```

The clean lockfile is acyclic, so `bw_cycles()` returns an empty list. The
diamond dependencies, ordered by how many packages need them, point at the
highest-risk shared packages:

```{r diamonds}
diamonds <- bw_diamonds(g)
head(diamonds[, c("package", "n_dependents")])
```

## Detect conflicts

`detect_conflicts()` returns a zero-row tibble when a library is healthy. The
clean fixture is healthy:

```{r conflicts-clean}
detect_conflicts(bw_example("clean.lock"))
```

The `conflicted.lock` fixture is not. It surfaces four problems: a missing
dependency, a hard version conflict, a version drift, and a cycle. The
`required_by` column is a list column, so convert it to a string before printing
as a `data.frame`:

```{r conflicts-bad}
conf <- detect_conflicts(bw_example("conflicted.lock"))
conf$required_by <- vapply(conf$required_by, paste, character(1), collapse = ", ")
as.data.frame(conf[, c("package", "type", "required_by", "satisfiable")])
```

Reading the rows: `zeta` is **missing** (needed by `alpha`); `rlang` is a
**version_conflict** (one package wants `>= 1.1.0`, another `<= 0.4.12`); `cli`
is a **version_drift** (a package wants `>= 99.0.0` but the pinned `cli` is
`3.6.2`); and `ouroboros`/`boros` form a **cycle**.

## Check system requirements

`check_sysreqs()` maps packages to the OS libraries they need on a given
platform. Passing `host_libraries` as a character vector makes presence
deterministic: a library counts as present only if it appears in that vector.
With an empty vector, nothing is installed, so every requirement is reported
missing:

```{r sysreqs}
sys <- check_sysreqs(
  bw_example("missing-sysreq.lock"),
  platform = "linux",
  host_libraries = character()
)
as.data.frame(sys[, c("package", "system_requirement", "status")])
```

Supplying, say, `host_libraries = c("libcurl", "openssl")` would flip the
matching rows to `status = "ok"`.

## Simulate the install

`simulate_install()` produces a dependency-first install order and predicts an
outcome for every package, propagating failures: anything that depends
(transitively) on a failing package is marked `blocked`.

```{r simulate}
plan <- simulate_install(bw_example("conflicted.lock"))
plan$feasible

as.data.frame(plan$predicted_failures[, c("package", "predicted_status", "reason")])
```

`feasible` is `FALSE`, and `predicted_failures` lists each package that will
fail or be blocked along with the reason -- `alpha` blocked by the missing
`zeta`, `rlang` and its dependents tripped by the version conflict, and the
`ouroboros`/`boros` cycle.

## One-call diagnosis

`diagnose()` runs the scan, graph, conflict, system-requirement and install
steps together and bundles the results. Its `summary()` method returns a compact
two-column health tibble:

```{r diagnose}
diag <- diagnose(
  bw_example("conflicted.lock"),
  platform = "linux",
  host_libraries = character()
)
summary(diag)
```

The individual pieces are available as `diag$library`, `diag$graph`,
`diag$conflicts`, `diag$sysreqs` and `diag$plan`.

From a diagnosis (or any path/library) you can also produce a shareable
artifact. `report(x, format = "quarto")` renders a self-contained HTML report and
returns its path, and `run_dashboard(x)` launches an interactive Shiny dashboard.
Both are omitted here because they render or open a browser; call them from an
interactive session:

```{r report, eval = FALSE}
# Renders inst/report to a standalone HTML file and returns the path
report(bw_example("conflicted.lock"), format = "quarto")

# Launches the interactive Shiny dashboard
run_dashboard(bw_example("conflicted.lock"))
```

## Gate an install with the pre-install hook

`bw_preinstall_check()` is the hook to wire into a `renv`/`pak` workflow or CI
step. It runs the full diagnosis against a lockfile and, with `strict = TRUE`,
aborts when problems exist -- failing the build before a single package is
fetched:

```{r hook, eval = FALSE}
# Returns a bw_diagnosis invisibly; aborts under strict = TRUE if problems exist
bw_preinstall_check("renv.lock", strict = TRUE)
```

That is the whole loop: scan a lockfile, inspect the graph and conflicts, check
system libraries, simulate the install, and gate on the result -- all before the
real install runs.
